From physical principles to ???

### 2016-03-09 2015CompleteDerivation/completeDerivation.tex
More work on composite state definition

Suppose we study the motion of a cannonball, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules scatter off its surface randomly, its trajectory is not greatly affect as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. In other words: the set of states must also be closed under non-deterministic. If the Brownian motion is not negligible, we do not end in a well defined position/momentum pair, even if we start from one. %Reworded this for Isaac

@@

%TODO: need more examples within explanation to make more intuitive

@@

\subsection{Homogeneous decomposable systems}
The notion that the system is decomposable means the states is equipped with a rule of composition that allows to write $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$: the composite system is the sum of its part. For example, the state of a ball is equal to the state of its left and right parts.

@@

\begin{defn}\label{reducible_state_space}
The state space $\mathcal{C}$ for a homogeneous decomposable system is an additive abelian (i.e. commutative) group.
\end{defn}

\begin{justification}
We claim $\mathcal{C}$ is an additive monoid. There exist a law of composition $+ : \mathcal{C} \times \mathcal{C} \rightarrow \mathcal{C}$ that takes two states and returns one that is the physical composition of the two. The domain and codomain match because the system is homogeneous. The law is commutative $\mathcal{c}_1 +\mathcal{c}_2 = \mathcal{c}_2+\mathcal{c}_1$ and associative $(\mathcal{c}_1 + \mathcal{c}_2) + \mathcal{c}_3 = \mathcal{c}_1 + (\mathcal{c}_2 + \mathcal{c}_3)$, as it does not matter in what order we physically compose the parts. There exist a unique zero element $\mathcal{c} + 0 = \mathcal{c}$ and it represent the physically empty state (i.e. no system).

We claim $\mathcal{C}$ is an additive group. As we want to describe changes during the evolution (i.e. $\hat{\mathcal{c}} = \mathcal{c} +\Delta \mathcal{c}$) we introduce an inverse $- : \mathcal{C} \rightarrow \mathcal{C}$ such that $\mathcal{c} + ( - \mathcal{c}) = 0 \forall \mathcal{c} \in \mathcal{C}$. Such inverse may introduce objects that do not properly represent a physical state, but a state change. A change of a physically distinguishable object is itself physically distinguishable therefore is physically well defined. We'll still call $\mathcal{C}$ the state space committing an abuse of terminology.
\end{justification}

Note that, due to \ref{statedef}, $\mathcal{C}$ is also a topological space.

@@

We will call a classical system one that is infinitesimally reducible. That is: it is made of arbitrarily small parts, which we call particles, and the state of the whole system is equivalent to the sum of the states of the parts. Given the state space $\mathcal{S}$ for the infinitesimal parts, a composite state $\mathcal{c}$ will tell us the amount of material for each possible particle state. That is, such states comes with a function $Q: \mathcal{S}\times\mathcal{C} \to \mathbbm{R}$ that returns the amount of material in the composite state $\mathcal{c}$ that is prepared in state $\mathcal{c}$. For example, given a certain configuration of gas, we can tell the amount that is at a specific point in place with a specific value of momentum. The notion that the parts are infinitesimal requires the co-domain of $Q$ to be continuous, as opposed to discrete. Note that the state space $\mathcal{S}$, instead, can be finite. For example, for a system composed of different tanks connected by pipes, the state of the composite system could be the overall distribution of water among the tanks. The amount in each tank is continuous (as we assume the water to be infinitesimally divisible) yet there are only a finite number of tanks the water can be placed.

As we combine the state of finite parts, the distributions over particle states sum (i.e. they combine linearly). This gives us the structure of a real vector space.

\begin{defn}\label{classical_vector space}
The state space $\mathcal{C}$ for a homogeneous classical (i.e. infinitesimally reducible) system is a vector space over $\mathbbm{R}$. It admits a basis isomorphic to the state space $\mathcal{S}$ of its particles (i.e. infinitesimal parts).
\end{defn}

\begin{justification}
The state space is an abelian group as the system is homogeneous and decomposable and \ref{reducible_state_space}.

We now consider the set of transformations $A$ that increase or decrease the amount of parts of the system by a constant. We claim $A$ is a field\footnote{Here field is intended in the abstract algebraic sense (a nonzero commutative division ring) which has no relationship to the field in the physics sense (a physical quantity with a value for each point in space).} isomorphic to $\mathbbm{R}$. Consider $a: \mathbbm{R} \rightarrow A$ the mapping between a number and the transformation that increases or decreases the size of the system by that number. This transformation exists: the system is infinitesimally decomposable and the amount can be changed continuously. Define on $A$ an addition $+: A \times A \rightarrow A$ and a multiplication $*: A \times A \rightarrow A$ such that $a(x) + a(y) = a(x+y)$ and $a(x) * a(y) = a(x*y)$, $x,y \in \mathbbm{R}$, so that the sum and product of the transformation is equal to the sum and product of their respective factors. $a$ is an isomorphism between $A$ and $\mathbbm{R}$ as fields.

We now claim that the state space $\mathcal{C}$ is a vector space over $\mathbbm{R}$. The abelian group $\mathcal{C}$ can be extended with the operations defined by $A$, as each element $a \in A$ is a map $a : \mathcal{C} \rightarrow \mathcal{C}$. The map has the following properties: $(a_1 + a_2) \mathcal{c} = a_1 \mathcal{c} + a_2 \mathcal{c}$, increasing the size of the system by the sum of two constant is the same as combining the separate increases, and $a (\mathcal{c}_1 + \mathcal{c}_2) = a \mathcal{c}_1 + a \mathcal{c}_2$, increasing the size of the total system is the same as the combination of the increased parts. $\mathcal{C}$ is a module over $A$, which is a field and isomorphic to $\mathbbm{R}$.

Consider now the injection map $\mathcal{e}: \mathcal{S} \hookrightarrow \mathcal{C}$ that for each particle state returns a composite state constituted only by a predetermined amount of particles in such state. The image of such injection is a basis for $\mathcal{C}$. We first claim that all elements are linearly independent. Consider $\sum\limits_{\mathcal{s}_i \in \mathcal{S}} a_i \mathcal{e}(\mathcal{s}_i)$, $a_i \in \mathbbm{R}$. This corresponds to a physical state with $a_i$ particles in each state $\mathcal{s}_i$. This is equal to the empty state if and only if $a_i=0 \forall i$. This holds even if $\mathcal{S}$ is not finite. We now claim they span. Consider $\mathcal{c} + \sum\limits_{\mathcal{s}_i \in \mathcal{S}} a_i \mathcal{e}(\mathcal{s}_i)$, $a_i \in \mathbbm{R}$. This can always be made equal to the empty state by setting $a_i$ to be the opposite of the amount of particles of $\mathcal{c}$ that are in state $\mathcal{s}_i$.
\end{justification}

>>>>

Suppose we study the motion of a cannonball, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules scatter off its surface randomly, its trajectory is not greatly affect as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. In other words: the set of states must also be closed under non-deterministic evolution. If the Brownian motion is not negligible, we do not end in a well defined position/momentum pair, even if we start from one. %Reworded this for Isaac

@@

@@

\subsection{Homogeneous decomposable systems}
The notion that a system is decomposable means the states is equipped with a rule of composition that allows to write $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$: the composite system is the sum of its part. For example, the state of a ball is equal to the state of its left and right parts.

@@

\begin{defn}\label{reducible_state_space}
The state space $\mathcal{C}$ for decomposable systems made of a homogeneous material is an additive abelian (i.e. commutative) group.
\end{defn}

\begin{justification}
We claim $\mathcal{C}$ is an additive monoid. There exist a law of composition $+ : \mathcal{C} \times \mathcal{C} \rightarrow \mathcal{C}$ that takes two states and returns one that is the physical composition of the two. The domain and codomain match because the material is homogeneous. The law is commutative $\mathcal{c}_1 +\mathcal{c}_2 = \mathcal{c}_2+\mathcal{c}_1$ and associative $(\mathcal{c}_1 + \mathcal{c}_2) + \mathcal{c}_3 = \mathcal{c}_1 + (\mathcal{c}_2 + \mathcal{c}_3)$, as it does not matter in what order we physically compose the parts. There exist a unique zero element $\mathcal{c} + 0 = \mathcal{c}$ and it represent the physically empty state (i.e. no amount of material).

We claim $\mathcal{C}$ is an additive group. As we want to describe changes during the evolution (i.e. $\hat{\mathcal{c}} = \mathcal{c} +\Delta \mathcal{c}$) we introduce an inverse $- : \mathcal{C} \rightarrow \mathcal{C}$ such that $\mathcal{c} + ( - \mathcal{c}) = 0 \forall \mathcal{c} \in \mathcal{C}$. Such inverse may introduce objects that do not properly represent a physical state, but a state change. A change of a physically distinguishable object is itself physically distinguishable therefore $\mathcal{C}$ is still a Hausdorff and second countable topological manifold as per \ref{statedef}. We'll still call $\mathcal{C}$ state space committing an abuse of terminology.
\end{justification}

@@

We will call a classical system one that is infinitesimally reducible. That is: it is made of arbitrarily small parts, which we call particles, and the state of the whole system is equivalent to the sum of the states of the parts. Given the state space $\mathcal{S}$ for the infinitesimal parts, a composite state $\mathcal{c}$ will tell us the amount of material for each possible particle state. That is, each state is fully identified by a function $a: \mathcal{S} \to \mathbbm{R}$ that returns the amount of material in the composite state $\mathcal{c}$ that is prepared in state $\mathcal{s}$. For example, given a certain configuration of gas, we can tell the amount that is at a specific point in space with a specific value of momentum. The notion that the parts are infinitesimal requires the co-domain of $a$ to be continuous, as opposed to discrete. The state space $\mathcal{S}$, instead, can be finite. For example, for a system composed of different tanks connected by pipes, the state of the composite system could be the overall distribution of water among the tanks. The amount in each tank is continuous (as we assume the water to be infinitesimally divisible) yet there are only a finite number of tanks the water can be placed into.

As we combine the state of finite parts, we sum the distributions over particle states $a(\mathcal{s})=a_1(\mathcal{s})+a_2(\mathcal{s})$: the amount of material in the composite state is the sum of the amount of material of the parts. This gives us the structure of a real vector space.

\begin{defn}\label{classical_vector space}
The state space $\mathcal{C}$ for a homogeneous classical (i.e. infinitesimally reducible) material is a vector space over $\mathbbm{R}$. It admits a basis isomorphic to the state space $\mathcal{S}$ of its particles (i.e. infinitesimal parts).
\end{defn}

\begin{justification}
The state space is an abelian group as the material is homogeneous and decomposable and \ref{reducible_state_space}.

We now consider the set of transformations $T$ that increase or decrease the amount of material in the system by a constant. We claim $T$ is a field\footnote{Here field is intended in the abstract algebraic sense (a nonzero commutative division ring) which has no relationship to the field in the physics sense (a physical quantity with a value for each point in space).} isomorphic to $\mathbbm{R}$. Consider $\tau: \mathbbm{R} \rightarrow T$ the mapping between a number and the transformation that increases or decreases the amount of material by that number. This transformation exists: the system is infinitesimally decomposable and the amount can be changed continuously. Define on $T$ an addition $+: T \times T \rightarrow T$ and a multiplication $*: T \times T \rightarrow T$ such that $\tau(a) + \tau(b) = \tau(a+b)$ and $\tau(a) * \tau(b) = \tau(a*b)$, $a,b \in \mathbbm{R}$, so that the sum and product of the transformation is equal to the sum and product of their respective factors. $\tau$ is an isomorphism between $T$ and $\mathbbm{R}$ as fields.

We now claim that the state space $\mathcal{C}$ is a vector space over $\mathbbm{R}$. The abelian group $\mathcal{C}$ can be extended with the operations defined by $T$, as each element $\tau \in T$ is a map $\tau : \mathcal{C} \rightarrow \mathcal{C}$. The map has the following properties: $(\tau_1 + \tau_2) \mathcal{c} = \tau_1 \mathcal{c} + \tau_2 \mathcal{c}$, increasing the amount of material by the sum of two constant is the same as combining the separate increases, and $\tau (\mathcal{c}_1 + \mathcal{c}_2) = \tau \mathcal{c}_1 + a \mathcal{c}_2$, increasing the amount of the total system is the same as the combination of the increased parts. $\mathcal{C}$ is a module over $T$, which is a field and isomorphic to $\mathbbm{R}$. $\mathcal{C}$ is (isomorphic to) a real vector space.

Consider now the injection map $\mathcal{e}: \mathcal{S} \hookrightarrow \mathcal{C}$ that for each particle state returns a composite state constituted only by a predetermined unit amount of material in such state. The image of such injection is a basis for $\mathcal{C}$. We first claim that all elements are linearly independent. Consider $\sum\limits_{\mathcal{s} \in \mathcal{S}} \tau(a(\mathcal{s})) \mathcal{e}(\mathcal{s})$, $a: \mathcal{S} \to \mathbbm{R}$. This corresponds to a physical state with $a(\mathcal{s})$ material in each state $\mathcal{s}$. This is equal to the empty state if and only if $a(\mathcal{s})=0 \forall \mathcal{s} \in \mathcal{S}$. We now claim they span. Consider $\mathcal{c} + \sum\limits_{\mathcal{s} \in \mathcal{S}} \tau(a(\mathcal{s})) \mathcal{e}(\mathcal{s})$. This can always be made equal to the empty state by setting $a(\mathcal{s})$ to be the opposite of the amount of material of $\mathcal{c}$ found in state $\mathcal{s}$. Both arguments hold even if $\mathcal{S}$ is an infinite or uncountable set, as they hold for each finite subset of $\mathcal{S}$ independently and therefore $\mathcal{C}=\prod\limits_{\mathcal{s} \in \mathcal{S}} span({\mathcal{e}(\mathcal{s})})$.
\end{justification}
